---
title: "A. Introduction to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A. Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(stats)
library(graphics)
```

# Essential _R_

A simple calculator

```{r}
1 + 1
```

'Vectors' as building blocks

```{r}
c(1, 2, 3)
c("January", "February", "March")
c(TRUE, FALSE)
```

Variables, missing values and 'factors'

```{r}
age <- c(27, NA, 32, 29)
gender <- factor(
    c("Female", "Male", "Non-binary", NA),
    levels = c("Female", "Male", "Non-binary")
)
```

Data structures to coordinate related vectors -- the `data.frame`

```{r}
df <- data.frame(
    age = c(27, NA, 32, 29),
    gender = gender
)
df
```

Functions

```{r}
rnorm(5)        # 5 random normal deviates
x <- rnorm(100) # 100 random normal deviates
hist(x)         # histogram, approximately normal
plot(density(x)) # a little more sophisticated?
```

'Vectorized' operations, e.g., element-wise addition without an
explicit 'for' loop


```{r}
y <- x + rnorm(100)
plot(y ~ x)
fit <- lm(y ~ x)
fit         # an R 'object' containing information about the
            # regression of y on x
abline(fit) # plot points and fitted regression line
anova(fit)  # statistical summary of linear regression
```

Write your own functions

```{r}
hello <- function(who) {
    paste("hello", who, "with", nchar(who), "letters in your name")
}
hello("Martin")
```

Iterate, usually with `lapply()` although `for()` is
available

```{r}
names <- c("Martin", "Thomas")
lapply(names, hello)
```

# Packages

Extend functionality of base _R_. Can be part of the 'base' distribution...

```{r}
## iterate over the numbers 1 through 8, 'sleeping' for 1 second
## each. Takes about 8 seconds...
system.time({
    lapply(1:8, function(i) Sys.sleep(1))
})

## sleep in parallel -- takes only 2 seconds
library(parallel)
cl <- makeCluster(4) # cluster of 4 workers
system.time({
    parLapply(cl, 1:8, function(i) Sys.sleep(1))
})
```

...or contributed, e.g., [dplyr][] to introduce the 'tidyverse'

```{r, message = FALSE}
library(dplyr)
```

A 'tibble' is like a 'data.frame', but more user-friendly

```{r}
tbl <- tibble(
    x = rnorm(100),
    y = x + rnorm(100)
)

## e.g., only displays the first 10 rows
tbl
```

The tidyverse makes use of 'pipes' `|>`. A pipe takes the left-hand
side and pass through to the right-hand side. Key [dplyr][] 'verbs'
can be piped together: `filter()` (rows), `select()` (columns),
`mutate()` (change values), `group_by()` (operate on groups of rows)


```{r}
tbl |>
    ## e.g., just rows with non-negative values of x and y
    filter(x > 0, y > 0) |>
    ## add a column
    mutate(distance_from_origin = sqrt(x^2 + y^2))
```


Another example: [ggplot2][] for visualization

```{r}
library(ggplot2)
ggplot(tbl) +
    aes(x, y) +                # use 'x' and 'y' columns for plotting...
    geom_point() +             # ...plot points...
    geom_smooth(method = "lm") # ...linear regresion
```

## Where do Packages Come From?

- [CRAN][]: *C*omprehensive *R* *A*rchive *N*etwork. More than 18,000
  packages. Some help from [CRAN Task Views][] in identifying relevant
  packages.

- [Bioconductor][]: More than 2100 packages relevant to
  high-throughput genomic analysis. _Vignettes_ are an important part
  of _Bioconductor_ packages.
  
Install packages once per _R_ installation, using
`BiocManager::install(<package-name>)` (CRAN or Bioconductor)

What about GitHub? Packages haven't been checked by a formal system,
so may have incomplete code, documentation, dependencies on other
packages, etc. Authors may not yet be committed to long-term
maintenance of their package.

[CRAN]: https://cran.r-project.org/web/packages/available_packages_by_name.html
[CRAN Task Views]: https://cran.r-project.org/web/views/
[Bioconductor]: https://bioconductor.org/packages
[dplyr]: https://cran.r-project.org/package=dplyr
[ggplot2]: https://cran.r-project.org/package=ggplot2

# Help & Vignettes

1. Help pages, e.g., `?lm`

2. Vignettes, e.g., 

    ```{r, eval = FALSE}
    vignette(package = "ggplot2")
    vignette("ggplot2-specs", "ggplot2")
    ```

3. Google, StackOverflow, etc...
